---
title: "NHS R Data Visualisation Library"
author: "NHS England Regional Collaboration utilising the foundations of a GGplot library created by Mike Perham"
date: "17/05/2024"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This library has been created to enable users to view examples of data visualisations utilising the NHS themes, using NHS public data sets. Users can load the required packages and create the sample data sets, and then choose which data visualisations they would like to run in their own environment.

The foundations of this library are sourced from a GGplot guide by Mike Perham. This has been expanded as a proof of concept for working collaboratively over regions to input and build upon, becoming the R Data Viz Library.

For more information about each data visualisation type, the below are recommended to review:

### The Data Visualisation Catalogue: <https://datavizcatalogue.com/index.html>

### The R Graph Gallery: <https://r-graph-gallery.com/>

### Top 50 GGplot2 Visualisations: <https://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>

# Install/load required packages

```{r LoadData, warning = FALSE, message = FALSE}
if (!require("pacman")) install.packages("pacman"); library(pacman)

pacman::p_load(Rcpp, tidyverse,dplyr,tidyr,
               ggplot2,ggthemes,ggtext,scales,
               png,ggalt,NHSRdatasets,onsr,shinycssloaders, plotly)

# install.packages('devtools')
#devtools::install_github('nhs-r-community/NHSRtheme')
# install.packages("remotes")
remotes::install_github("rOpenSci/fingertipsR",
                        build_vignettes = TRUE,
                        dependencies = "suggests",
                        build = F)
```

# Load sample data

All of the examples in this document use A&E dummy data from the NHSRdatasets package for NHS reporting, fingertips data for Public Health and ONS data for population data. These give us broad datasets that can be used for different data visualisation types.

More information on these packages can be found here:

### NHSRdatasets:

<https://github.com/nhs-r-community/NHSRdatasets>

<https://nhs-r-community.github.io/NHSRdatasets/>

```{r , warning = FALSE}
#Load initial dataset and clean up
Attends <- NHSRdatasets::ae_attendances %>%
  filter(org_code == "RXQ"|org_code=="RTH"|org_code=="RHW"|
           org_code == "RTK"|org_code == "RA2") %>%
  filter(type ==1) %>%
  select(-c(3,5))
```

### Fingertips:

<https://github.com/ropensci/fingertipsR>

The below gets population by sex.

```{r , warning = FALSE, eval = FALSE}
# Load fingertipsR package
library(fingertipsR)

# Get available profiles in fingertips
profiles_data <- profiles()
print(profiles_data)

# Find profiles related to population
population_profiles <- profiles_data %>%
  filter(grepl("population", ProfileName, ignore.case = TRUE))
print(population_profiles)

# Search for indicators related to population structure, DomainID was found by viewing the population_profiles and selecting a DomainID
population_indicators <- indicator_metadata(DomainID = "1938133081") #Replace with DomainID from above

# Get the relevant indicator of the measure from your list of population_indicators
indicator_id <- 92708  # Replace with the actual indicator ID from the results above

#Get list of area types
area_types_data <- area_types()

# Get the data for the specific indicator
population_data <- fingertips_data(IndicatorID = indicator_id,AreaTypeID = "15") #Note for Area Types you can select all, but it will take a long time. 15 is England.

# Filter data to include only relevant columns and non-NA values
population_data_filtered <- population_data %>%
  filter(!is.na(Age), !is.na(Value)) %>%
  select(AreaName, Sex, Age, Value)

# Adjust the values for plotting (male values negative for pyramid structure)
population_data_filtered <- population_data_filtered %>%
  mutate(Value = ifelse(Sex == "Male", -Value, Value)) %>%
  filter(Age != "All ages") %>%
  filter(Sex != "Persons")

# Convert the Age column to a factor and specify the levels in the desired order
population_data_filtered$Age <- factor(population_data_filtered$Age, levels = c("0-4 yrs", "5-9 yrs", "10-14 yrs", "15-19 yrs", "20-24 yrs", "25-29 yrs", "30-34 yrs", "35-39 yrs", "40-44 yrs", "45-49 yrs", "50-54 yrs", "55-59 yrs", "60-64 yrs", "65-69 yrs", "70-74 yrs", "75-79 yrs", "80-84 yrs", "85-89 yrs","90+ yrs"))
```

### ONS:

[https://medium.com/\@VickyCrockett1/how-do-you-get-data-into-r-from-the-ons-c860043fef8c](https://medium.com/@VickyCrockett1/how-do-you-get-data-into-r-from-the-ons-c860043fef8c){.uri}

The next step loads the data and performs some simple filtering steps.

# NHSR Theme

All of the examples in this document use dummy data from the NHSRdatasets package (more information on this package can be found here: <https://github.com/nhs-r-community/NHSRtheme>). As the package is not in CRAN, you need to use devtools to load the package from github.

```{r , warning = FALSE, echo = FALSE, warning = FALSE}

remotes::install_github("nhs-r-community/NHSRtheme")

NHSRtheme::get_nhs_colours()

```

# Data Over Time 

## Basic line chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}
#Filter initial dataset
line_df <- Attends %>%
  filter(org_code=="RXQ")

#Make plot
ggplot(line_df, aes(x = period, y = attendances)) +
  geom_line(colour = "#005EB8", size = 1.5) +
  scale_y_continuous(labels = comma) +
  labs(title="Type 1 attendances - Bucks Healthcare",
       subtitle = "April 2016 to March 2019",
       y = "Attendances",
       x = "Month") +
  expand_limits(y = 0)

```

### Plotly

NEY TO ADD IN

## Multiple line chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}
#Filter initial dataset
multiple_line_df <- Attends %>%
  filter(org_code == "RXQ" | org_code=="RTH") 

#Make plot
ggplot(multiple_line_df,
         aes(x = period, y = attendances, colour = org_code)) +
  geom_line(size = 1) +
  geom_point() +
  scale_colour_manual(values = c("#005EB8", "#41B6E6")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Type 1 attendances - Bucks Healthcare vs Royal Berkshire",
    subtitle = "April 2016 to March 2019",
    y = "Attendances",
    x = "Month"
  ) +
  expand_limits(y = 0) +
  theme(legend.title = element_blank())

```

### Plotly

NEY TO ADD IN

## SPC Charts

NEY TO ADD IN

# Comparisons

## Simple bar chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}
#Filter initial dataset
bar_df <- Attends %>%
  filter(period == "2019-03-01")

#Make plot
bar <- ggplot(bar_df, aes(x = org_code, y = attendances)) +
  geom_bar(stat = "identity",
           position = "identity",
           fill = "#005EB8") +
  geom_hline(yintercept = 0,
             size = 1,
             colour = "#333333") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Type 1 attendances",
    subtitle = "March 2019",
    y = "Attendances",
    x = "Provider Code"
  )

plot(bar)

```

**Add labels**

The code below adds labels to your simple bar chart.

```{r , warning = FALSE}
#Filter initial dataset
bar + geom_text(aes(label = scales::comma(attendances)), vjust =2, color= "White")

```

### Plotly

NEY TO ADD IN

## Grouped bar chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}
#Filter initial dataset
grouped_bar_df <- Attends %>%
  filter(period == "2017-03-01" | period == "2019-03-01") %>%
  select(c(1:3))

#Make plot
ggplot(grouped_bar_df,
       aes(
         x = org_code,
         y = attendances,
         fill = as.factor(period)
       )) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 0,
             size = 1,
             colour = "#333333") +
  scale_y_continuous(labels = comma) +
  #NHSRtheme::scale_fill_nhs('blues')+
  labs(
    title = "Attendances have increased in all providers other than Bucks Healthcare",
    subtitle = "March 2017 vs March 2019",
    y = "Attendances",
    x = "Provider Code"
  ) +
  theme(legend.title = element_blank())

```

### Plotly

NEY TO ADD IN

## Stacked bar chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE, message = FALSE}
AttendsAll <- NHSRdatasets::ae_attendances %>%
  filter(
    org_code == "RXQ" | org_code == "RTH" | org_code == "RHW" |
      org_code == "RTK" | org_code == "RA2"
  ) %>%
  filter(period == '2017-03-01')

ggplot(AttendsAll, aes(fill = type, y = attendances, x = org_code)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_y_continuous(labels = comma) +
  labs(title = "A&E attendances by department type - March 2017",
       y = "Attendances",
       x = "Provider Code") +
  theme(legend.title = element_blank())

```

### Plotly

NEY TO ADD IN

## Percent stacked bar chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}
AttendsAll <- NHSRdatasets::ae_attendances %>%
  filter(
    org_code == "RXQ" | org_code == "RTH" | org_code == "RHW" |
      org_code == "RTK" | org_code == "RA2"
  ) %>%
  filter(period == '2017-03-01')

ggplot(AttendsAll, aes(fill = type, y = attendances, x = org_code)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = percent) +
  labs(title = "A&E attendances by department type - March 2017",
       y = "Attendances",
       x = "Provider Code") +
  theme(legend.title = element_blank())

```

### Plotly

NEY TO ADD IN

## Bubble chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}

#For this example we are filtering on 5 organisations, type 1 activity & excluding column 3 from the dataframe.
Attends <- NHSRdatasets::ae_attendances %>%
  filter(org_code == "RXQ"|org_code=="RTH"|org_code=="RHW"|
           org_code == "RTK"|org_code == "RA2")

  # Summarise data for type 1 attendances
  type_1_summary <- Attends %>%
  filter(type == 1) %>%
  group_by(org_code, period) %>%
  summarise(type_1_attendances = sum(attendances, na.rm = TRUE)) %>%
  ungroup()

# Summarise data for non-type 1 attendances
type_other_summary <- Attends %>%
  filter(type == 3|type== 2) %>%
  group_by(org_code, period) %>%
  summarise(type_other_attendances = sum(attendances, na.rm = TRUE)) %>%
  ungroup()

# Summarise total attendances and total admissions
total_summary <- Attends %>%
  group_by(org_code, period) %>%
  summarise(
    total_attendances = sum(attendances, na.rm = TRUE),
    total_breaches = sum(breaches, na.rm = TRUE),
    total_admissions = sum(admissions, na.rm = TRUE)
  ) %>%
  ungroup()

# Merge the summaries into a single data frame
final_summary <- total_summary %>%
  left_join(type_1_summary, by = c("org_code", "period")) %>%
  left_join(type_other_summary, by = c("org_code", "period"))

# Replace NA values with 0 for type_1_attendances and type_3_attendances
final_summary <- final_summary %>%
  mutate(
    type_1_attendances = replace_na(type_1_attendances, 0),
    type_other_attendances = replace_na(type_other_attendances, 0)
  )

# Add percentage columns
final_summary <- final_summary %>%
  mutate(
    perc_admissions_attendances = (total_admissions / total_attendances) * 100,
    perc_type1_attendances_total = (type_1_attendances / total_attendances) * 100,
    perc_breaches_attendances = (total_breaches / total_attendances) * 100
  )

# Filter initial dataset
bubble_df <- final_summary

# Calculate size for bubble chart (proportional to type 1 attendances)
bubble_df <- bubble_df %>%
  mutate(size = perc_type1_attendances_total / max(perc_type1_attendances_total) * 100)

# Create bubble chart
ggplot(bubble_df, aes(x = perc_admissions_attendances, y = perc_breaches_attendances, size = size, color = size)) +
  geom_point(alpha = 0.5) +
  scale_size_continuous(name = "Proportion of type 1") +
  #NHSRtheme::scale_fill_nhs('blues', name = "Proportion of type 1") +
  labs(title = "Bubble Chart of % 4 Hour Breaches vs % converted to admission with % Attendances Type 1 Size",
       x = "Conversion Rate", y = "% 4 Hour Breaches") +
  theme(legend.position = "right")
```

### Plotly

NEY TO ADD IN

## Population Pyramid chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}
# Plot the population pyramid
ggplot(population_data_filtered, aes(x = Age, y = Value, fill = Sex)) +
  geom_bar(stat = "identity", position = "identity") +
  coord_flip() +
  scale_y_continuous(labels = function(x) comma(abs(x))) +
  labs(title = "Population Age Profile by Gender",
       x = "Age Group",
       y = "Population Count",
       fill = "Gender") +
  NHSRtheme::scale_fill_nhs("blues")
```

### Plotly

NEY TO ADD IN

## Radar chart

Insert here

## Radial chart

Insert here

## Radial Column chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}

# Create a radial column chart
ggplot(Attends, aes(x = reorder(org_code, -attendances), y = attendances)) +
  geom_col(width = 0.5, fill = "skyblue") +
  coord_polar(start = 0) +
   #NHSRtheme::scale_fill_nhs("blues") +
  scale_y_continuous(labels = function(x) comma(abs(x))) +
  labs(title = "Attendances per Trust",
       x = NULL, y = NULL)
```

### Plotly

NEY TO ADD IN

## Span chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}
# Summarize the data
summary_data <- Attends %>%
  group_by(org_code) %>%
  summarise(min_attendance = min(attendances),
            max_attendance = max(attendances))

# Create the horizontal bar range chart
ggplot(summary_data, aes(y = org_code)) +
  geom_linerange(aes(xmin = min_attendance, xmax = max_attendance), color = "blue", size = 1.5) +
  labs(title = "Range of Type 1 Attendances by Trust between 2019 and 2023",
       x = "Number of Attendances",
       y = "Organisation Code") +
   NHSRtheme::scale_fill_nhs("blues")
```

### Plotly

NEY TO ADD IN

## Stacked area chart {.tabset .tabset-fade}

### ggplot

```{r , warning = FALSE}

Attends <- NHSRdatasets::ae_attendances %>%
  filter(org_code == "RXQ"|org_code=="RTH"|org_code=="RHW"|
           org_code == "RTK"|org_code == "RA2")

# Plot the stacked area chart
ggplot(Attends, aes(x = period, fill = org_code)) +
  geom_area(stat = "count") +  
  NHSRtheme::scale_fill_nhs("blues")

```

### Plotly

NEY TO ADD IN

```{r , warning = FALSE}

# Plot the stacked area chart 

blues <- c("#00A9CE", "#41B6E6", "#0072CE",   "#005EB8","#003087" )

Attends <- NHSRdatasets::ae_attendances %>%
  filter(org_code == "RXQ"|org_code=="RTH"|org_code=="RHW"|
           org_code == "RTK"|org_code == "RA2") |>
  # sort by unique code in reverse so the chart is stacked in opposite alphabet order
  mutate(org_code = factor(org_code, levels = rev(sort(unique(org_code)))))

plot_ly(data = Attends |> group_by(period, org_code) |>  summarise(count = n()) |> ungroup(),
        x = ~period,
        y = ~count,
        color = ~org_code,
        colors = blues,
        type = 'scatter',
        mode = 'none',
        stackgroup = 'one',
        line = list(colors = blues),
        stackgroup = 'one') |>
# solid fill colours to replicate the ggplot style
  style(traces = 1, fillcolor = "#00A9CE") |>
  style(traces = 2, fillcolor = "#41B6E6") |>
  style(traces = 3, fillcolor = "#0072CE") |>
  style(traces = 4, fillcolor = "#005EB8") |>
  style(traces = 5, fillcolor = "#003087") |>
  layout(legend = list(title = list(text = "org_code")),
         xaxis = list(tickformat = "%Y",
                      dtick = "M12"))


```



# Creating multiple charts for the same measure

## Faceted chart

Use `facet_wrap()` to create multiple charts split by subgroup in data. You can use `ncol =` or `nrow` to specify number of rows or columns. For example, `facet_wrap(~org_code, nrow=1)` to put all charts in a single row.

```{r , warning = FALSE}
ggplot(Attends, aes(x = period, y = attendances)) +
  geom_line(colour = "#005EB8", size = 1) +
  facet_wrap(~org_code)+
  labs(title="Type 1 attendances",
       subtitle = "April 2016 to March 2019") +
  expand_limits(y = 0)

```

# Range

## Box & Whisker chart

```{r , warning = FALSE}
g <- ggplot(Attends, aes(org_code, attendances))
g + geom_boxplot(varwidth=T, fill="light blue") + 
    labs(title="A&E Attendances", 
         subtitle="Distribution by Trust",
         caption="Source: A&E Monthly Stats",
         x="Trust",
         y="A&E attendances")
```       

## Bullet chart

Insert here

## Candlestick chart

Insert here

## Error bars chart

Insert here

## Funnel Plot

NEY TO ADD IN

## Gantt chart

Insert here

## Kagi chart

Insert here

## Span chart

Insert here

## Violin chart

```{r, warning = FALSE}

#Violin Plot
ggplot(Attends, aes(org_code, attendances)) + geom_violin() + 
  labs(title="A&E Attendances", subtitle="Range by Trust", caption="Source: A&E Monthly Statistics", x="Trust", y="Attendances") + scale_fill_brewer(palette="Blues") + theme_classic()
```

## Dumbell chart {.tabset .tabset-fade}

### ggplot

The `ggtext` package can be used to add colour to titles or subtitles. You need to ensure that you use it with `theme(plot.subtitle = element_markdown(hjust = 0, size = 12))` otherwise it will not work.

```{r, warning = FALSE}
#Prepare data
dumbbell_df <- NHSRdatasets::ae_attendances %>%
  filter(type ==1) %>%
  select(-c(3,6)) %>%
  filter(period == "2017-03-01" | period =="2019-03-01") %>%
  mutate(period =as.numeric(format(period,'%Y'))) %>%
  mutate(period = as.character(period)) %>%
    mutate(performance = 1- (breaches/attendances)) %>%
  select(c(1:2,5)) %>%
  spread(period, performance) %>%
  mutate(gap = `2019` - `2017`) %>%
  arrange(desc(gap)) %>%
  head(10)

#Make plot
dumbell <- ggplot(dumbbell_df, aes(x = `2017`, xend = `2019`, y = reorder(org_code, gap), group = org_code)) + 
  geom_dumbbell(colour = "#dddddd",
                size = 3,
                colour_x = "#41B6E6",
                colour_xend = "#005EB8") +
  scale_x_continuous(labels = scales::percent_format(accuracy=1))+
  geom_vline(xintercept = 0.95, size = 1, colour="#333333", linetype = "dashed") +
  labs(title = "Performance improved for all providers",
    subtitle = "<span style='color: #41B6E6;'>March 2017 <span><span style='color: black;'> vs <span><span style='color: #005EB8;'>March 2019<span>") +
  xlab("4 hour performance") +
  ylab("Org code") +
 theme(plot.subtitle = element_markdown(hjust = 0, size = 12))+ theme(legend.position = "none")

plot(dumbell)
```

**Adding annotations**

You can use `geom_label` to add annotations to existing plots or you can add line in when creating ggplot.

```{r, warning = FALSE}

dumbell + geom_label(aes(x = 0.9, y = "R1K",label = "Standard"), 
                           hjust = -0.5, 
                           vjust = -0.1, 
                           colour = "#555555",
                           label.size = NA, 
                           family="Arial", 
                           size = 4)

```

### Plotly

NEY TO ADD IN

# Movement or Flow

## Sankey Diagram

NEY TO ADD IN
